---
output: pdf_document
---

# Appendix: Full R Code and Output

```{r setup, include=FALSE}
library(tidyverse)
library(ggrepel)
```

```{r, include=FALSE}
# Load data
df <- read.csv("PlayerStats.csv")

# Clean data
df <- df[, !names(df) %in% c("X", "X.1")]
df$team <- trimws(df$team)
```

## ARI vs UTA summary table
```{r}
players_summary <- df %>%
  filter(team %in% c("ARI", "UTA")) %>%
  group_by(player_name, team) %>%
  summarise(
    avg_ppg = sum(points) / sum(gp),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = team,
    values_from = avg_ppg
  ) %>%
  rename(
    arizona_ppg = ARI,
    utah_ppg = UTA
  )
```

## ARI vs UTA Plot
```{r}
ggplot(players_summary, aes(x = arizona_ppg, y = utah_ppg, label = player_name)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  geom_point(size = 3, color = "#8B1A1A") +
  ggrepel::geom_label_repel(size = 3) +
  labs(
    title = "Former Coyotes: PPG Before and After Relocation",
    x = "Points Per Game in Arizona",
    y = "Points Per Game in Utah"
  ) +
  theme_minimal()
```

## Crouse analysis
```{r}
crouse_df <- df %>%
  filter(player_name == "Lawson Crouse") %>%
  arrange(season)
crouse_df <- crouse_df %>%
  mutate(season = factor(season, levels = unique(season)))

ggplot(crouse_df, aes(x = season, y = ppg, group = 1, color = team)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("ARI" = "#8B1A1A", "UTA" = "#5DADE2")) +
  labs(
    title = "Lawson Crouse: Career PPG",
    x = "Season",
    y = "Points Per Game",
    color = "Team"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Player overview bar chart
```{r}
knitr::kable(players_summary, 
             caption = "Average PPG: Arizona vs Utah Era",
             digits = 3)
bar_df <- players_summary %>%
  pivot_longer(
    cols = c(arizona_ppg, utah_ppg),
    names_to = "team",
    values_to = "ppg"
  ) %>%
  mutate(team = recode(team,
                       arizona_ppg = "Arizona",
                       utah_ppg = "Utah"))

ggplot(bar_df, aes(x = player_name, y = ppg, fill = team)) +
  geom_col(position = "dodge") +
  labs(
    title = "Arizona vs Utah: Points Per Game by Player",
    x = "Player",
    y = "Points Per Game",
    fill = ""
  ) +
  scale_fill_manual(values = c("Arizona" = "#8B1A1A", "Utah" = "#5DADE2")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## Percent change (ARI --> UTAH)
```{r}
plot1_df <- players_summary %>%
  drop_na(arizona_ppg, utah_ppg) %>%
  mutate(
    pct_change = (utah_ppg - arizona_ppg) / arizona_ppg * 100
  ) %>%
  arrange(pct_change)

ggplot(plot1_df, aes(x = reorder(player_name, pct_change), y = pct_change)) +
  geom_col(fill = "#5DADE2") +
  coord_flip() +
  labs(
    title = "Percent Change in Scoring After Relocation (Arizona to Utah)",
    x = "Player",
    y = "Percent Change in PPG"
  ) +
  theme_minimal()
```

## Creating years_in_system variable

```{r}
joining_year <- c(
  "Clayton Keller"    = 2016,
  "Nick Schmaltz"     = 2019,
  "Lawson Crouse"     = 2016,
  "Michael Carcone"   = 2021,
  "Logan Cooley"      = 2023,
  "Sean Durzi"        = 2023,
  "Barrett Hayton"    = 2019,
  "Alexander Kerfoot" = 2023,
  "Michael Kesselring"= 2022,
  "Matias Maccelli"   = 2021,
  "Jack McBain"       = 2021
)
```
```{r}
# Extract the start year from the season string
df <- df %>%
  mutate(
    season_start = as.numeric(substr(season, 1, 4)),
    years_in_system = season_start - joining_year[player_name] + 1
  )

df %>%
  select(player_name, season, years_in_system) %>%
  arrange(player_name, season) %>%
  head(50)
```

## Functions
```{r}
# Activation functions
sigmoid <- function(z) 1 / (1 + exp(-z))
sigmoid_deriv <- function(z) {
  s <- sigmoid(z)
  s * (1 - s)
}
tanh_deriv <- function(z) 1 - tanh(z)^2

# Forward pass
forward_pass <- function(X, W1, b1, W2, b2,
                         activation = "tanh") {
  Z1 <- X %*% t(W1) +
    matrix(b1, nrow(X), length(b1), byrow = TRUE)
  if (activation == "sigmoid") {
    A1 <- sigmoid(Z1)
  } else {
    A1 <- tanh(Z1)
  }
  f <- A1 %*% W2 + b2
  list(Z1 = Z1, A1 = A1, f = f)
}

# Loss functions
mse_loss <- function(y, y_hat) mean((y - y_hat)^2)
r_squared <- function(y, y_hat) {
  1 - sum((y - y_hat)^2) / sum((y - mean(y))^2)
}

# Train function
train_nn <- function(X, y, K = 10, lr = 0.01,
                     epochs = 500, lambda = 0.01,
                     activation = "tanh",
                     batch_size = 10,
                     seed = 123) {
  set.seed(seed)
  n <- nrow(X)
  p <- ncol(X)
  W1 <- matrix(rnorm(K * p, sd = 0.1), K, p)
  b1 <- rep(0, K)
  W2 <- rnorm(K, sd = 0.1)
  b2 <- 0
  loss_history <- numeric(epochs)
  for (epoch in 1:epochs) {
    idx <- sample(n)
    X_shuf <- X[idx, , drop = FALSE]
    y_shuf <- y[idx]
    for (start in seq(1, n, by = batch_size)) {
      end <- min(start + batch_size - 1, n)
      X_batch <- X_shuf[start:end, , drop = FALSE]
      y_batch <- y_shuf[start:end]
      m <- nrow(X_batch)
      fp <- forward_pass(X_batch, W1, b1, W2, b2, activation)
      residual <- fp$f - y_batch
      dW2 <- (t(fp$A1) %*% residual) / m + 2 * lambda * W2
      db2 <- mean(residual)
      delta <- residual %*% t(W2)
      if (activation == "sigmoid") {
        delta <- delta * sigmoid_deriv(fp$Z1)
      } else {
        delta <- delta * tanh_deriv(fp$Z1)
      }
      dW1 <- (t(delta) %*% X_batch) / m + 2 * lambda * W1
      db1 <- colMeans(delta)
      W1 <- W1 - lr * dW1
      b1 <- b1 - lr * db1
      W2 <- W2 - lr * dW2
      b2 <- b2 - lr * db2
    }
    fp_full <- forward_pass(X, W1, b1, W2, b2, activation)
    loss_history[epoch] <- mse_loss(y, fp_full$f)
  }
  list(W1 = W1, b1 = b1, W2 = W2, b2 = b2,
       loss_history = loss_history)
}
```

## Simulated Data: Why Linear Regression Fails
```{r}
set.seed(42)
n_sim <- 100
x_sim <- seq(-pi, pi, length.out = n_sim)
y_sim <- sin(x_sim) + rnorm(n_sim, sd = 0.2)

sim_df <- data.frame(x = x_sim, y = y_sim)

ggplot(sim_df, aes(x = x, y = y)) +
  geom_point(alpha = 0.6, color = "gray40") +
  geom_smooth(method = "lm", color = "#8B1A1A",
              se = FALSE, linetype = "dashed") +
  labs(
    title = "Linear Regression: Sine Wave",
    subtitle = "A straight line cannot capture nonlinear structure",
    x = "X", y = "Y"
  ) +
  theme_minimal()

X_sim <- matrix(x_sim, ncol = 1)
X_sim_scaled <- scale(X_sim)
```

## Effect of Number of Hidden Units (K)
```{r}
k_values <- c(2, 5, 20)
results_sim <- list()

for (k in k_values) {
  model_sim <- train_nn(
    X_sim_scaled, y_sim,
    K = k,
    lr = 0.05,
    epochs = 1000,
    lambda = 0.001,
    activation = "tanh",
    batch_size = 20
  )
  fp_sim <- forward_pass(X_sim_scaled, model_sim$W1,
                         model_sim$b1, model_sim$W2, model_sim$b2)
  results_sim[[paste0("K=", k)]] <- as.numeric(fp_sim$f)
}
```
```{r}
sim_plot_df <- data.frame(
  x = rep(x_sim, 3),
  y_true = rep(y_sim, 3),
  y_pred = c(results_sim[["K=2"]],
             results_sim[["K=5"]],
             results_sim[["K=20"]]),
  K = rep(c("K = 2 units", "K = 5 units", "K = 20 units"),
          each = n_sim)
)

ggplot(sim_plot_df, aes(x = x)) +
  geom_point(aes(y = y_true), alpha = 0.3, color = "gray50") +
  geom_line(aes(y = y_pred, color = K), linewidth = 1.2) +
  labs(
    title = "Neural Network Fitting a Sine Wave: Effect of K Units",
    subtitle = "More hidden units = more flexibility",
    x = "X", y = "Y",
    color = "Hidden Units"
  ) +
  theme_minimal() +
  facet_wrap(~K)
```

## Sigmoid vs Tanh Activation Functions
```{r}
activations <- c("sigmoid", "tanh")
act_results <- list()

for (act in activations) {
  model_act <- train_nn(
    X_sim_scaled, y_sim,
    K = 5,
    lr = 0.05,
    epochs = 1000,
    lambda = 0.001,
    activation = act,
    batch_size = 20
  )
  fp_act <- forward_pass(X_sim_scaled, model_act$W1,
                         model_act$b1, model_act$W2, model_act$b2,
                         activation = act)
  act_results[[act]] <- as.numeric(fp_act$f)
}

act_plot_df <- data.frame(
  x = rep(x_sim, 2),
  y_true = rep(y_sim, 2),
  y_pred = c(act_results[["sigmoid"]],
             act_results[["tanh"]]),
  activation = rep(c("Sigmoid", "Tanh"), each = n_sim)
)

ggplot(act_plot_df, aes(x = x)) +
  geom_point(aes(y = y_true), alpha = 0.3, color = "gray50") +
  geom_line(aes(y = y_pred, color = activation), linewidth = 1.2) +
  scale_color_manual(values = c("Sigmoid" = "#8B1A1A", 
                                "Tanh" = "#5DADE2")) +
  labs(
    title = "Sigmoid vs Tanh Activation Function",
    subtitle = "K = 5 units, comparing two activation functions",
    x = "X", y = "Y",
    color = "Activation"
  ) +
  theme_minimal()
```

## Effect of L2 Regularization
```{r}
lambdas <- c(0.0001, 0.01, 0.5)
reg_results <- list()

for (lam in lambdas) {
  model_reg <- train_nn(
    X_sim_scaled, y_sim,
    K = 20,
    lr = 0.05,
    epochs = 1000,
    lambda = lam,
    activation = "tanh",
    batch_size = 20
  )
  fp_reg <- forward_pass(X_sim_scaled, model_reg$W1,
                         model_reg$b1, model_reg$W2, model_reg$b2)
  reg_results[[paste0("lambda=", lam)]] <- as.numeric(fp_reg$f)
}

reg_plot_df <- data.frame(
  x = rep(x_sim, 3),
  y_true = rep(y_sim, 3),
  y_pred = c(reg_results[["lambda=1e-04"]],
             reg_results[["lambda=0.01"]],
             reg_results[["lambda=0.5"]]),
  lambda = rep(c("lambda = 0.0001 (low)", 
                 "lambda = 0.01 (balanced)", 
                 "lambda = 0.5 (high)"), each = n_sim)
)

ggplot(reg_plot_df, aes(x = x)) +
  geom_point(aes(y = y_true), alpha = 0.3, color = "gray50") +
  geom_line(aes(y = y_pred, color = lambda), linewidth = 1.2) +
  labs(
    title = "Effect of L2 Regularization on Neural Network Fit",
    subtitle = "K = 20 units: effect of varying lambda",
    x = "X", y = "Y",
    color = "Regularization"
  ) +
  theme_minimal() +
  facet_wrap(~lambda)
```

## Effect of Learning Rate

```{r}
learning_rates <- c(0.001, 0.01, 0.1)
lr_results <- list()

for (lr in learning_rates) {
  model_lr <- train_nn(
    X_sim_scaled, y_sim,
    K = 5,
    lr = lr,
    epochs = 1000,
    lambda = 0.001,
    activation = "tanh",
    batch_size = 20
  )
  fp_lr <- forward_pass(X_sim_scaled, model_lr$W1,
                        model_lr$b1, model_lr$W2, model_lr$b2)
  lr_results[[paste0("lr=", lr)]] <- as.numeric(fp_lr$f)
}

lr_plot_df <- data.frame(
  x = rep(x_sim, 3),
  y_true = rep(y_sim, 3),
  y_pred = c(lr_results[["lr=0.001"]],
             lr_results[["lr=0.01"]],
             lr_results[["lr=0.1"]]),
  lr = rep(c("lr = 0.001 (slow)", 
             "lr = 0.01 (balanced)", 
             "lr = 0.1 (fast)"), each = n_sim)
)

ggplot(lr_plot_df, aes(x = x)) +
  geom_point(aes(y = y_true), alpha = 0.3, color = "gray50") +
  geom_line(aes(y = y_pred, color = lr), linewidth = 1.2) +
  labs(
    title = "Effect of Learning Rate on Neural Network Fit",
    subtitle = "K = 5 units, varying learning rate",
    x = "X", y = "Y",
    color = "Learning Rate"
  ) +
  theme_minimal() +
  facet_wrap(~lr)
```

## Neural network code

```{r}
# ------------------------------------------------------------
# 1. Prepare Data
# ------------------------------------------------------------

df <- df %>%
  mutate(season_start = as.numeric(substr(season, 1, 4)))

nn_df <- df %>%
  arrange(player_name, season_start) %>%
  group_by(player_name) %>%
  
  mutate(prior_ppg_raw = lag(ppg)) %>%

  mutate(
    prior_ppg = ifelse(is.na(prior_ppg_raw), ppg, prior_ppg_raw)
  ) %>%
  
  ungroup() %>%
  select(player_name, team, season, ppg, age,
         years_in_system, avg_toi, prior_ppg)

# ------------------------------------------------------------
# 2. Train/Test Split
# ------------------------------------------------------------

set.seed(123)
train_idx <- sample(seq_len(nrow(nn_df)),
                    size = 0.8 * nrow(nn_df))

train_df <- nn_df[train_idx, ]
test_df  <- nn_df[-train_idx, ]

X_train <- train_df %>%
  select(age, years_in_system, avg_toi, prior_ppg) %>%
  as.matrix()

y_train <- train_df$ppg

X_test <- test_df %>%
  select(age, years_in_system, avg_toi, prior_ppg) %>%
  as.matrix()

y_test <- test_df$ppg

# ------------------------------------------------------------
# 3. Scale Features
# ------------------------------------------------------------

X_scaled_train <- scale(X_train)
X_mean <- attr(X_scaled_train, "scaled:center")
X_sd   <- attr(X_scaled_train, "scaled:scale")

X_scaled_test <- scale(X_test, center = X_mean, scale = X_sd)

# ------------------------------------------------------------
# 4. Train Model
# ------------------------------------------------------------

model <- train_nn(
  X_scaled_train, y_train,
  K = 10,
  lr = 0.01,
  epochs = 500,
  lambda = 0.01,
  activation = "tanh",
  batch_size = 10
)

# Plot training loss curve
data.frame(
  epoch = 1:length(model$loss_history),
  loss = model$loss_history
) %>%
  ggplot(aes(x = epoch, y = loss)) +
  geom_line(color = "#8B1A1A", linewidth = 1) +
  labs(
    title = "Training Loss Over Time",
    subtitle = "MSE decreasing across 500 epochs",
    x = "Epoch",
    y = "MSE Loss"
  ) +
  theme_minimal()

# ------------------------------------------------------------
# 5. Predictions
# ------------------------------------------------------------

fp_train <- forward_pass(X_scaled_train, model$W1, model$b1,
                         model$W2, model$b2)

fp_test <- forward_pass(X_scaled_test, model$W1, model$b1,
                        model$W2, model$b2)

train_df$predicted_ppg <- as.numeric(fp_train$f)
test_df$predicted_ppg  <- as.numeric(fp_test$f)

train_df$residual <- train_df$ppg - train_df$predicted_ppg
test_df$residual  <- test_df$ppg - test_df$predicted_ppg

nn_results <- bind_rows(train_df, test_df)

# ------------------------------------------------------------
# 6. Evaluation
# ------------------------------------------------------------

cat("Train MSE:", mse_loss(y_train, train_df$predicted_ppg), "\n")
cat("Test MSE:", mse_loss(y_test, test_df$predicted_ppg), "\n")

cat("Train R2:", r_squared(y_train, train_df$predicted_ppg), "\n")
cat("Test R2:", r_squared(y_test, test_df$predicted_ppg), "\n")

# ------------------------------------------------------------
# 7. Plots
# ------------------------------------------------------------

ggplot(nn_results,
       aes(x = predicted_ppg,
           y = ppg,
           color = team)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", color = "gray") +
  geom_point(size = 3) +
  scale_color_manual(values = c(
    "ARI" = "#8B1A1A",
    "UTA" = "#5DADE2"
  )) +
  labs(
    title = "Actual vs Expected PPG (Neural Network Model)",
    x = "Predicted PPG",
    y = "Actual PPG",
    color = "Team"
  ) +
  theme_minimal()

# ------------------------------------------------------------
# Plot: color = team, shape = player
# ------------------------------------------------------------

focus_players <- c("Clayton Keller", "Lawson Crouse", "Logan Cooley")

plot_df <- nn_results %>%
  filter(player_name %in% focus_players)

ggplot(plot_df,
       aes(x = predicted_ppg,
           y = ppg,
           color = team,
           shape = player_name)) +

  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed",
              color = "gray") +

  geom_point(size = 4) +

  # Team colors (ARI vs UTA)
  scale_color_manual(values = c(
    "ARI" = "#8B1A1A",   # deep red
    "UTA" = "#5DADE2"    # light blue
  )) +

  # Player shapes
  scale_shape_manual(values = c(
    "Clayton Keller" = 16,
    "Lawson Crouse" = 17,
    "Logan Cooley" = 15
  )) +

  labs(
    title = "Actual vs Expected PPG: Key Players",
    x = "Expected PPG (Neural Net)",
    y = "Actual PPG",
    color = "Team",
    shape = "Player"
  ) +

  theme_minimal()
```

## Model Residuals by Team

```{r}
ggplot(nn_results, aes(x = team, y = residual, fill = team)) +
  geom_boxplot(alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  scale_fill_manual(values = c(
    "ARI" = "#8B1A1A",
    "UTA" = "#5DADE2"
  )) +
  labs(
    title = "Model Residuals by Team",
    subtitle = "Positive residual = outperformed expectations",
    x = "Team",
    y = "Residual (Actual - Predicted PPG)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

